Computing Entropies With Nested Sampling

### 2017-08-14 paper/response.tex

%\abstract{\noindent Abstract}

% Need this after the abstract
\setlength{\parindent}{0pt}
\setlength{\parskip}{8pt}

\section*{Response to John Skilling's comments}

{\quote
This is an interesting paper, which addresses how to compute entropy values directly, without explicit
evaluation of the underlying probability distribution. There are a few places where I feel that presentation
could be improved, which I list below. Nevertheless, I should note two particular points.

1. Definition of entropy}

I agree with Skilling that the fundamental definition of entropy involves
a base measure $q(x)$, and that relative entropies or KL divergences are the
appropriate objects to be working with. However, since (as he points out)
the algorithm more naturally involves the averages of log probabilities,
I have decided to leave most of the discussion in terms of Shannon entropies.
However, I have added extra discussion to the introduction, about how using
the Shannon entropy is equivalent to a uniform $q$, and cautioning that
this should be kept in mind. I have also added some text after the definition
of the log-volume of a ball of radius $r$, which clarifies where the implicit
assumption of a uniform $q(x)$ comes into the procedure.

{\quote
2. Possible future work
There is an infelicity at the heart of the author’s approach. The little target balls of radius $L$ need to be
small in order to avoid undue smudging of the pointwise values being estimated. Yet, as L shrinks, the
estimates lose accuracy because of the square-root-N variability of nested-sampling volumes. So the user
will need to compromise, and it’s possible that future work may discover how to ameliorate this.}

This is an excellent point, and something I have worried about. I have added
some discussion of this in Section 4 (search for the paragraph containing
`trade-off'). I think the current algorithm will still work well when the
thing we want the entropy of is low-dimensional, even if it's some function of
an underlying high-dimensional quantity.

{\quote
I suggest that applying the author’s method directly to the Kullback-Leibler form (equations 11 and
12) may offer a better way forward.}

It was not obvious to me how this would work (except perhaps from using a
distance function defined in the coordinate system in which $q$ is uniform).
I will think about it, and present any future work in this way if it is
sufficiently neat.

{\quote 
Near the end of page 9, the author argues against trying that
because “the estimate for a given dataset may be biassed”. Possibly so, though I would argue that the
dominant uncertainty with nested sampling is statistical, with numerical biasses (of sensible quantities)
being considerably less. Anyway, it’s not entirely clear in these nonlinear problems whether bias matters
much. I have in mind the Bayesian evidence Z, where an unbiassed estimate would give a seriously
biassed estimate of the logarithm log Z (which is what matters for Bayes factors), and vice versa.
The advantage of aiming at KL is that the statistic can be computed “for free” parasitically upon
a nested sampling run. In fact, I myself use apparent convergence of KL as the most reliable general
termination criterion I have yet found. (Hint to author: it converges quicker if it’s evaluated as the sum
$KL =
X \delta L
Z
log
L
Z$
over likelihood L, rather than using the corresponding weighted sum over prior mass shells $\delta X$ that’s
better for getting Z and the posterior.)}

It is not the bias that I was particularly worried about. I have added a
comment about the bias potentially being negligible in many applications.
What I was more concerned about was being able to integrate over nuisance
parameters as well. I can't see a way to get the prior-to-posterior
KL divergence {\em of nuisance parameters only} using standard NS, but I can
get it using my approach.

{\quote
To the editors: I am not suggesting that there’s anything wrong with the author’s proposals. We
advance by trying things, and even if something does become superseded — which this may not — it
1still gives helpful perspective.}

I agree. In fact, this paper has gone through a few distinct stages where I
started seeing better ways of doing certain things. It's possible that it
will be superseded quickly, but I do think it's an advance. I'd like to get
people thinking about Nested Sampling in different scenarios, i.e., where
pi and L are 'any distribution' and 'any scalar function' rather than necessarily
being prior and likelihood. We might find applications to things other than
data analysis and statistical mechanics by doing that.

{\quote
Minor improvements
1. Wrong English in last sentence of 1.1 (page 2).}

{\quote
2. “were” not “where” at beginning of 2 (page 2) — use brain not spellchecker.}

{\quote
3. I found the notation a bit confusing. $H(\theta | d)$ looks like a function of the posterior $\theta$’s at specific
data d, whereas it actually means the average hHi for unknown d. I agree that it’s unnecessarily fussy
to write the formally correct H(P (· | ·)) for H as a functional of the posterior as a function of arbitrary
arguments which readers are supposed to remember are parameters and data. However, angle brackets,
$hH(\theta | d)i$ or $H(\theta | d)i d$ would have helped me by giving a useful clue. They might also let the author
avoid the idiosyncratic $H(\theta | d = d 0 )$ by writing $H(\theta | d 0 )$ without the angle brackets for H at the specific
d 0 .
There are various overloaded definitions of H according to type of argument (see equations 8, 9, 10).
Introducing an explicit common format
X X
H(argument) $\equiv$
p($\theta$, d) log p(argument)
would bring these together and incidentally help to stress the basic point that the paper is addressing
averages of logarithms.}

I thought about this for a while, and decided not to change the notation much.
The reason is that my notation is closer to what is used in introductory
information theory texts. One usually sees things such as $H(Y|X=x)$
and $H(Y|X)$, where the latter is the expected value of the former (averaged
over $X$). This is the standard definition of conditional entropy.
I think my writing makes it clear what I mean.
However, in my idiosyncratic $H(\theta | d = d_0)$ I decided to change the
subscript 0 to a subscript obs for observed.

{\quote
4. At the top of page 5, I find it a bit nicer to initialise nested sampling with X 0 = 1 rather than
X 1 = t 1 . That also helps with the evaluation...}

Unfortunately we have competing tastes again at this point.
I don't like to include $X_0$ because there's no zeroth iteration of Nested Sampling.

{\quote
5. I realise that L is the active constraint during iterations, but I wonder if readers might get confused
by the use of symbol L as radius (mnemonic length?) whereas the standard usage has been L = likelihood.}

Yes, $L$ was for length, and it was unintentional that I had $L$ for both
likelihood and length. I've changed this to $r$ for radius throughout.

{\quote
6. Equation 27 is messy in form and mistaken in typography. My memorable formula for
the volume of a hypersphere is}

Fixed, and I've replaced the volume formula with the given (easier to read and
remember) suggestion.

{\quote
7. Near the bottom of page 7 (Algorithm 1), I was confused by
b
h $\leftarrow$ b
h + + [k/N ]
Why double +? Why the space between them? Do square brackets mean integer rounding? If so, should
it be [k]/N ?}

Sorry about the confusion. I was trying to represent the operation of
appending one new element (a depth estimate)
to an existing list, which in Python is {\tt +}
and in Haskell (my current flavour of the month language) is {\tt ++}. The square brackets were meant to represent
the wrapping of a value within a list (to make a list of length one),
since an append function is probably a function of two lists.
All of this was too obscure. I have replaced it with
\begin{align}
\widehat{\mathbf{h}} \leftarrow \textnormal{append}\big(\widehat{\mathbf{h}}, k/N\big)
\end{align}
which is hopefully more understandable.

{\quote
8. In the caption of Figure 1, “around the red point” or “at the red ball” might be better than “at
the red point”.}

Changed.

{\quote
9. At equation 29,
1/100
should be
1/100.01
to conform to the 4-figure accuracy elsewhere.}

That was actually a mistake (I forgot to factor in the non-uniform prior),
rather than overzealous rounding. Thanks for noticing it.

{\quote
10. Below equation (45), I think ``$\ln(2 \times 10 ^{-5} )$'' would better be explained as ``ln(2L) = -10.820''.}

Fixed. It now says $\ln (2r)$.

{\quote
11. In the last paragraph of the text (page 19), there is a correct statement about how “to increase
one’s confidence in the results, [without getting] a guarantee”. It might be worth pointing out that there
can never be such a guarantee for any method that relies on pointwise evaluations, because there will
always be some risk of a dominant spike being missed.}

I've added such a sentence.

{\quote
12. In the references, my search for ref. 20 gave me “Statistics and Computing, 2017, 27, 219–236”.}

Somehow my bibliography file had become corrupted. I have fixed this.

@@

That's why I started thinking about this problem...

{\quote
1. The author presents a methodology for computing entropies, and other information-theoretic quantities, using Skilling's nested sampling algorithm.  It should be noted that Skilling himself advocates the estimation of the "entropy", I believe that it is the KL-divergence, as the estimate of the uncertainty in the estimated evidence.  This should be mentioned, and possibly discussed within the context of this present paper.}

I have mentioned that NS computes the prior-to-posterior KL divergence, and
that this has both an interesting interpretation and is used as a termination
criterion. This was added to the first paragraph of Section 3.

{\quote
2. With regard to the algorithm to compute tSorry about the confusion. I was trying to represent the operation of
appending one new element (a depth estimate)
to an existing list, which in Python is {\tt +}
and in Haskell is {\tt ++}. The square brackets were meant to represent
the wrapping of a value within a list (to make a list of length one),
since an append function is probably a function of two lists.
All of this was too obscure. I have replaced it with
\begin{align}
\widehat{\mathbf{h}} \leftarrow \textnormal{append}\big(\widehat{\mathbf{h}}, k/N\big)
\end{align}
which is hopefully more understandable.he depth, it is not clear what '+ +' refers to in the step in which the latest estimate is appended to the results.  Also at that step, do the square brackets refer to an array (of one element)?  Or do they refer to some kind of rounding, which I doubt.}

### 2017-08-14 paper/ms.tex
Finalish changes

If we actually want the differential entropy
\begin{align}
H(\x) &= -\int p(\x) \log p(\x) \, d\x
\end{align}
we can use the fact that density equals mass divided
by volume. Assume $r$ is small, so that
\begin{align}
P(d(\x; \xref) < r \,|\, \xref)
    &\approx
    p(\xref) \int_{d(\x; \xref) < r} \, d\x\\
    &= p(\xref) \, V
\end{align}
where $V$ is the volume of the region where $d(\x; \xref) < r$.
Then $H(\x) = \left< \depth \right>_{p(\x)} + \log V(r)$.
{\revision Reducing $r$ towards zero ought to lead to
a more accurate result because the probability density at a point will
be more accurately approximated by the probability of a small region containing
the point, divided by the volume of that region.
On the other hand, smaller values of $r$ will lead to more Nested Sampling
iterations required, and hence less accuracy in the depth estimates.
Thus, there is a an awkward trade-off involved in choosing the value of $r$,
which will need to be investigated more thoroughly in the future.}

@@

As the algorithm lets us calculate the entropy of any distribution which can
be sampled by MCMC, there are several options. The mutual information
can be written in various ways, such as the following.
\begin{align}
I(\x; \d) &= H(\x) + H(\d) - H(\x, \d) \\
I(\x; \d) &= H(\d) - H(\d|\x) \\
I(\x; \d) &= H(\x) - H(\x|\d) \\
I(\x; \d) &= \int p(\d) \, D_{\rm KL}\left(p(\x|\d) \,||\, p(\x)\right) \, d\d\\
I(\x; \d) &= D_{\rm KL}\left(p(\x, \d) \,||\, p(\x)p(\d)\right).
\end{align}
In a Bayesian problem where $\x$ is an unknown parameter and $\d$ is data, the
first and second formulas would be costly to compute, because the
high dimensional probability distribution for the dataset would require a
large number of NS iterations to compute $H(\d)$ and $H(\x, \d)$.
The fourth involves the Kullback-Leibler divergence from the prior to the
posterior, averaged over all possible datasets. This is straightforward to
approximate with standard Nested Sampling, but the estimate for a given dataset
may be biased. This method also would not work if we want $I(\x; \d)$ for
a single parameter or a subset of them, rather than for all of the parameters.
Therefore, I adopted the third formula as the best way to compute $I(\x; \d)$.
In the example, $H(\x)$ is available analytically, and I use the algorithm to
obtain $H(\x | \d)$ as follows:
\begin{align}
H(\x | \d) &= \int p(\d_{\rm ref}) H(\x | \d=\d_{\rm ref}) \, d\d_{\rm ref}.
\end{align}
This is an expected value over datasets, of the entropy of the posterior given
each dataset. I ran the algorithm by generating reference particles from
$p(\x, \d)$, i.e., generating parameter values from the prior, along with
corresponding simulated datasets. Given each simulated dataset, I used NS
to compute an unbiased estimate of $H(\x | \d=\d_{\rm ref})$, by evolving
a set of particles (initially from the posterior) towards the reference particle.
The role of the ``prior'' in the NS is actually taken by the posterior, and the
``likelihood'' was minus a distance function defined in terms of the
parameter(s).

>>>>

If we actually want the differential entropy
\begin{align}
H(\x) &= -\int p(\x) \log p(\x) \, d\x
\end{align}
we can use the fact that density equals mass divided
by volume. Assume $r$ is small, so that
\begin{align}
P(d(\x; \xref) < r \,|\, \xref)
    &\approx
    p(\xref) \int_{d(\x; \xref) < r} \, d\x\\
    &= p(\xref) \, V
\end{align}
where $V$ is the volume of the region where $d(\x; \xref) < r$.
Then $H(\x) = \left< \depth \right>_{p(\x)} + \log V(r)$.
{\revision It is important to remember that the definition of
volume here is in terms of the coordinates $\x$, and that a change
of coordinates (or a change of the definition of the distance function)
would result in a different value of the
differential entropy.}

{\revision Reducing $r$ towards zero ought to lead to
a more accurate result because the probability density at a point will
be more accurately approximated by the probability of a small region containing
the point, divided by the volume of that region.
On the other hand, smaller values of $r$ will lead to more Nested Sampling
iterations required, and hence less accuracy in the depth estimates.
Thus, there is a an awkward trade-off involved in choosing the value of $r$,
which will need to be investigated more thoroughly in the future.
However, the algorithm as it currently stands ought to be useful for calculating
the entropies of any low-to-moderate dimensional functions of an
underlying high-dimensional space.}

@@

As the algorithm lets us calculate the entropy of any distribution which can
be sampled by MCMC, there are several options. The mutual information
can be written in various ways, such as the following.
\begin{align}
I(\x; \d) &= H(\x) + H(\d) - H(\x, \d) \\
I(\x; \d) &= H(\d) - H(\d|\x) \\
I(\x; \d) &= H(\x) - H(\x|\d) \\
I(\x; \d) &= \int p(\d) \, D_{\rm KL}\left(p(\x|\d) \,||\, p(\x)\right) \, d\d\\
I(\x; \d) &= D_{\rm KL}\left(p(\x, \d) \,||\, p(\x)p(\d)\right).
\end{align}
In a Bayesian problem where $\x$ is an unknown parameter and $\d$ is data, the
first and second formulas would be costly to compute, because the
high dimensional probability distribution for the dataset would require a
large number of NS iterations to compute $H(\d)$ and $H(\x, \d)$.
The fourth involves the Kullback-Leibler divergence from the prior to the
posterior, averaged over all possible datasets. This is straightforward to
approximate with standard Nested Sampling, but the estimate for a given dataset
may be biased. {\revision It is quite plausible that the bias is negligible
compared to the variance from averaging over many datasets. This is probably
true in many situations}.

However, more significantly, the above method
would not work if we want $I(\x; \d)$ for
a single parameter or a subset of them, rather than for all of the parameters.
{\revision
The prior-to-posterior KL divergence measured by standard NS is the
amount learned about {\em all} parameters, including nuisance parameters}.
Therefore, I adopted the third formula as the best way to compute $I(\x; \d)$.
In the sinusoidal example, $H(\x)$ is available analytically, and I use the algorithm to
obtain $H(\x | \d)$ as follows:
\begin{align}
H(\x | \d) &= \int p(\d_{\rm ref}) H(\x | \d=\d_{\rm ref}) \, d\d_{\rm ref}.
\end{align}
This is an expected value over datasets, of the entropy of the posterior given
each dataset. I ran the algorithm by generating reference particles from
$p(\x, \d)$, i.e., generating parameter values from the prior, along with
corresponding simulated datasets. Given each simulated dataset, I used NS
to compute an unbiased estimate of $H(\x | \d=\d_{\rm ref})$, by evolving
a set of particles (initially from the posterior) towards the reference particle.
The role of the ``prior'' in the NS is actually taken by the posterior, and the
``likelihood'' was minus a distance function defined in terms of the
parameter(s) of interest.
